Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex
fixed things in introduction

In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

\begin{example}
Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=2$, $\pi(3)=4$, $\pi(4)=1$, and $\pi(5)=3$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
\end{example}

\begin{remark}
There are two notations for permutations, `one-line' notation and `cycle' notation. We give definitions of both below. Unless stated otherwise, we will use one-line notation in this paper.
\end{remark}

@@

\begin{definition}
In cycle notation we identify cycles in the permutation and write them as such. Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$. Then the cycle notation for $\pi$ is $$(1 \ \pi(1) \ \pi(\pi(1)) \cdots \pi^{k-1}(1))(\alpha_1 \ \pi(\alpha_1) \ \pi(\pi(\alpha_1)) \cdots)\cdots(\alpha_\ell \ \pi(\alpha_\ell) \ \pi(\pi(\alpha_\ell)) \cdots).$$ Let $k$ be the smallest natural number for which $\pi^k(1)=1$. Then $\pi^{k-1}(1)$ is the last entry in the cycle before it is closed. If there are elements of $\{1,2,\cdots, n\}$ that are not in the cycle then let $\alpha_1$ be the smallest such element and continue the process until all elements are accounted for. 
\end{definition}

\begin{remark}
Note, cycle notation is not unique. 
\end{remark}

\begin{example}
Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, and $\pi(5)=3$. Then in one-line notation $\pi=54123$. In cycle notation $\pi=(153)(24)$. 
\end{example}

@@

%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$. There are two conventions commonly used to write elements of $S_n$. The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g. 43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.

%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$. For example, the cycle $(14)(23)$ sends $12345$ to $43215$.

@@

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}
\section{Terms in the sequence $s_n(312)$}
\label{S3}

@@

\section{Terms in the sequence $s_n(321)$}

Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.

>>>>

In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ for $\pi$ an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.

\begin{theorem*}
For $\pi$ an element of $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
\end{theorem*}

Of particular interest is the fact that two entirely distinct methods of proof had to be used to prove this theorem for all six permutations. In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

\begin{remark}
There are two notations for permutations, `one-line' notation and `cycle' notation. We define both below. Unless stated otherwise, we will use one-line notation in this paper.
\end{remark}

\begin{example}
Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, $\pi(5)=3$, and $\pi(6)=6$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
\end{example}

@@

\begin{example}
Using $\pi$ from the previous example, in one-line notation $\pi=541236$.
\end{example}

\begin{definition}
A cycle $(a_1\cdots a_n)$ with $1\le a_i\le n$ acts as a permutation on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
\end{definition}

A more formal definition of cycle notation is especially cumbersome, so we proceed with the example above, a permutation $\pi$ with $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, $\pi(5)=3$, and $\pi(6)=6$. To write $\pi$ in cycle notation we start with 1 and identify the cycle $1\mapsto 5\mapsto 3\mapsto 1$. Once we come back to 1 we close the cycle and write $(135)$. We keep identifying cycles until every number is accounted for. So $2\mapsto 4\mapsto 2$ gives us $(24)$ and $6\mapsto 6$ gives us $(6)$. Finally we can write $\pi$ as the composition of functions $(135)(24)(6)$. Note, cycle notation is not necessarily unique.

@@

@@

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}
\section{Terms in the sequence $s_n(312)$}
\label{312}

@@

\section{Terms in the sequence $s_n(321)$}
\label{321}
Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.
